A self-consistency check for unitary propagation of Hawking quanta

### 2016-09-09 Draft_BH_shells.tex
Small changes made to appendix
intro update

The black hole information paradox \cite{Haw76a} was sharpened to show a self-inconsistency among the following three widely-believed statements \cite{AMPS}
\footnote{Different physicists may believe in some of those more strongly than the others, but very few would have argued against any single statement without the explicitly conflict in the information paradox.}:
\begin{itemize}
\item {\bf Unitary evaporation:} A black hole will totally evaporate; the formation and evaporation of a black hole is described by an asymptotic observer as a unitary $S$-matrix, whose size is the exponential of the black hole's Bekenstein entropy.
\item {\bf General relativity:} The collapse-Schwarzschid geometry given by general relativity is a valid description to the spacetime everywhere except for near the singularity.
Therefore, there is no drama while crossing the horizon.
\item {\bf Local quantum field theory:} Away from the singularity, we can apply local quantum field theory (QFT) which describes microscopic steps of evaporation. 
For large black holes, it manifests as a unitary propagation of every near-horizon mode into every asymptotic Hawking quanta.
\end{itemize}
Basically, Statement Three demands that a near horizon mode is related to an asymptotic Hawking quantum by a unitary transformation, therefore carrying the same qubit of information. 
However statement 1 and 2 demand this unique qubit to be maximally entangled with two distinct objects, violating monogamy. 
This conflict has inspired many different proposals to modify one of the above three statements. 
For example, information loss or remnant \cite{Bek94} modifies Statement One; 
firewall \cite{BraPir09,AMPS} or ER=EPR \cite{MalSus13} modifies Statement Two;
various proposals of nonlocal effects near the horizon \cite{Gid12,DodSil15,OsuPag16} or causal patch complementarity \cite{HuiYan13,IlgYan13,LowTho14}
\footnote{Independent of the particular realization, causal-patch complementarity basically claims that QFT breaks down if one applies it to a region that does not fit into any observer's causal patch.} 
modifies Statement Three.

@@

Before presenting our argument, we will first explain its relation to some existing ideas.
The recent paper by Osuga and Page \cite{OsuPag16} provides an abstract framework of how modifying Statement Three resolves the paradox.
Basically, the near horizon mode is thermal as demanded by Statement Two.
But it propagates out in a non-unitary process, gaining the appropriate information to restore a unitary $S$-matrix as demanded by Statement One. 
For such abstract model to work in practice, such non-unitary propagation should be understood as interacting with a hidden system.
Here hidden simply means something that is ignorant to a na\"ive application of local QFT.
The obvious candidate of such hidden system is the geometry.

One proposal along this line of thoughts was discussed in \cite{NomVar12}. 
The idea was that the black hole acquires a large uncertainty during the evaporation process. 
If we treat the superposition of different classical geometries (black holes with different locations/momenta) as a quantum system, it may qualify as the hidden system that is responsible for the non-unitary propagation of Hawking quanta. 
Unfortunately, such proposal can be defeated by counting entropy.
In order for a non-unitary process to restore the asymptotic $S$-matrix, this hidden system must store the required information, which is comparable to the mount of microscopic information carried by the black hole.
However, the black-hole-no-hair theorem states that two black holes with identical macroscopic parameters (mass, angular momentum, change) are isometric to each other. 
This limits the number of different classical geometries to be parametrized by macroscopic parameters only.
As a result, the Hilbert space of superposed classical geometries has far less dimensions than necessary to carry the required amount of information.\footnote{We thank Raphael Bousso for private communication.}
Thus, even if a propagating Hawking quantum does interact with this hidden system, it cannot reproduce a unitary $S$-matrix; therefore such modification of Statement Three alone is insufficient to resolve the paradox.

The recent development on black-hole-soft-hairs \cite{HawPer16} hinted a way to revive this idea.
Basically, it was realized that two isometric (regions of) geometries should not always be considered as the same state.
If one maps the geometries with congruences of geodesics, two isometric regions with different maps are actually different states.
That is because changes in the mapping geodesics are measurable memory effects, and a consistent physical theory must be able to distinguish them.
This realization dramatically increases the number of different states encoded in classical geometries.
A tentative counting of state on a black hole horizon yields exactly the required number such that a modification to Statement Three can indeed resolve the paradox.

Before diving into details, we should explain why we focus on the last statement.
That is because local QFT simply assumes {\bf subsystem unitarity}.
Quantum mechanical evolution of a full system is by-definition unitary. 
However in practice, we are not able to describe everything by quantum mechanics all together, so we describe only subsystems. 
Namely, we assume that the full system can be separated into a ``classical background'' and a ``quantum subsystem''. 
After such artificial separation, since we only use quantum mechanics to describe the quantum subsystem, it always appears to be unitary.
\footnote{Even a time-dependent Hamiltonian leads to an apparently unitary evolution under this assumption. That is obviously not a physical fact since a unitary subsystem should be determined by its own initial state, not co-determined by initial state plus an arbitrary time-dependence.} 
However, such unitarity should not be taken as a physical fact unless one specifically checks the nature of interactions with the background.

>>>>

The black hole information paradox \cite{Haw76a} was sharpened to show a self-inconsistency among the following three widely-believed statements \cite{AMPS}
\footnote{Different physicists may believe in some of those more strongly than the others, but very few would have argued against any single statement without the explicitly conflict in the information paradox.}:
\begin{itemize}
\item {\bf Unitary evaporation:} A black hole will totally evaporate; the formation and evaporation of a black hole is described by an asymptotic observer as a unitary $S$-matrix, whose size is the exponential of the black hole's Bekenstein entropy.
\item {\bf General relativity:} The collapse-Schwarzschid geometry given by general relativity is a valid description to the spacetime everywhere except for near the singularity.
Therefore, there is no drama while crossing the horizon.
\item {\bf Local quantum field theory:} Away from the singularity, we can apply local quantum field theory (QFT) which describes microscopic steps of evaporation. 
For large black holes, it manifests as a unitary propagation of every near-horizon mode into every asymptotic Hawking quanta.
\end{itemize}
Basically, Statement Three demands that a near horizon mode is related to an asymptotic Hawking quantum by a unitary transformation, therefore carrying the same qubit of information. 
However Statement One and Two demand this unique qubit to be maximally entangled with two distinct objects, violating monogamy. 
This conflict has inspired many different proposals to modify one of the above three statements. 
For example, information loss or remnant \cite{Bek94} modifies Statement One; 
firewall \cite{BraPir09,AMPS} or ER=EPR \cite{MalSus13} modifies Statement Two;
various proposals of nonlocal effects near the horizon \cite{Gid12,DodSil15,OsuPag16} or causal patch complementarity \cite{HuiYan13,IlgYan13,LowTho14}
\footnote{Independent of the particular realization, causal-patch complementarity basically claims that QFT breaks down if one applies it to a region that does not fit into any observer's causal patch.} 
modifies Statement Three.

@@

Before presenting our argument, we will first explain its relation to some existing ideas.
The recent paper by Osuga and Page \cite{OsuPag16} provides an abstract framework of how modifying Statement Three resolves the paradox.
Basically, the near horizon mode is thermal as demanded by Statement Two.
But it propagates out in a non-unitary process, gaining the appropriate information to restore a unitary $S$-matrix as demanded by Statement One. 
For such abstract model to work in practice, this non-unitary propagation should be understood as interaction with a hidden system.
Here hidden simply means something that is ignorant to a na\"ive application of local QFT.
The obvious candidate of such hidden system is the geometry.

One proposal along this line of thoughts was discussed in \cite{NomVar12}. 
The idea was that the black hole acquires a large uncertainty during the evaporation process. 
If we treat the superposition of different classical geometries as a quantum system, it may qualify as the hidden system that is responsible for the non-unitary propagation of Hawking quanta. 
Unfortunately, such proposal can be defeated by counting entropy.
In order for such non-unitary process to restore the asymptotic $S$-matrix, the hidden system must store the required information, which is comparable to the mount of microscopic information carried by the black hole.
However, the black-hole-no-hair theorem states that two black holes with identical macroscopic parameters (mass, momentum, angular momentum, charge) are isometric to each other. 
This limits the number of different classical geometries to be parametrized by macroscopic parameters only.
As a result, the Hilbert space of superposed classical geometries has far less dimensions than necessary to carry the required amount of information.\footnote{We thank Raphael Bousso for private communication.}
Thus, even if a propagating Hawking quantum does interact with this hidden system, it cannot reproduce a unitary $S$-matrix; therefore such modification of Statement Three alone is insufficient to resolve the paradox.

The recent development on black-hole-soft-hairs \cite{HawPer16} hinted a way to revive this idea.
Basically, it was realized that two isometric (regions of) geometries should not always be considered as the same state.
If one maps the geometries with congruences of geodesics, two isometric regions with different maps are actually different states.
That is because changes in the mapping geodesics are measurable memory effects, and a consistent physical theory must be able to distinguish them.
This realization dramatically increases the number of different states encoded in classical geometries.
A tentative counting of state on a black hole horizon yields exactly the required number such that a modification to Statement Three can indeed resolve the paradox \cite{HawPer16}.

In this paper, we still take another step that is complementary to the above progress. It is exciting to know that modifying Statement Three can in-principle resolve the paradox. 
However, it remains a great mystery of why should local QFT break down in low curvature regions. 
If we just demand that it breaks down in order to resolve the information paradox, then there is no reason why this solution is preferred over modifying Statement One or Two. 
We would like argue that the application of QFT to a unitary propagation from a near horizon mode to an asymptotic Hawking quantum is already self-inconsistent. 
Thus it is only natural to modify Statement Three.

The starting point of our argument involves no new physics at all.
We simply recognize that local QFT implies a standard assumption of {\bf subsystem unitarity}.
Quantum mechanical evolution of a full system is by-definition unitary. 
However in practice, we are not able to describe everything by quantum mechanics all together, so we describe only subsystems. 
Namely, we assume that the full system can be separated into a ``classical background'' and a ``quantum subsystem''. 
After such artificial separation, since we only use quantum mechanics to describe the quantum subsystem, it always appears to be unitary.
\footnote{Even a time-dependent Hamiltonian leads to an apparently unitary evolution under this assumption. That is obviously not a physical fact since a unitary subsystem should be determined by its own initial state, not co-determined by initial state plus an arbitrary time-dependence.} 
However, such unitarity should not be taken as a physical fact unless one specifically checks the nature of interactions with the background.
